Summer 2025 ToDo

### 2025-07-08 Relations/Paper/Formalization.tex
Implications and summertodo minor changes update
Changes made on meeting 8th July

\apcomment{When SMseq comes up, also point to section 5, and the need for classical logic in going from
sequential versions to inductive ones.  IOW, elaborate the summary below. }

The following is a summary of the differences in definition we have used as mentioned in Section \ref{sec:Definitions}.
\begin{itemize}
    \item $\SN$ : We define this via the notion of accessibility rather than the non-existence of infinite reduction sequences. Both definitions articulate normalization as well-foundedness of the converse relation. However, different notions of well-foundedness are used in the two definitions. The differences are highlighted in Sections~\ref{sec:Well-foundedness}.
     See Section \ref{sec:Well-foundedness} for further discussion.
    \item $\NPe$ : We use the equivalent $\NP$ definition.
    \item $\Inc$ : We use the more general property $\RP$.
\end{itemize}

As mentioned, the $\Inc$ property can impose a computationally demanding requirement. However, $\Inc$ is used only in Theorem 1.2.3
and the only crucial aspect of $\Inc$ needed for the proof is that no $R$-increasing
sequence can have an upper bound.
We establish that the equivalent properties $\RP$ and $\RPm$ are sufficient for replacing $\Inc$ in the theorem (see Subsection \ref{subsec:theorems}),
which led us to examine the condition on $s (i)$ appearing inside $\RP$ more closely. This condition provides a generalization of the notion of a normal form,
which we explore in detail in Section \ref{sec:Hierarchy}.

@@

\begin{proof}
    See \texttt{module Theorem-1-2-2} in \texttt{Theorems.agda}, \\
    \verb|CR→NP : R isCR → R isNP₌|\\
    \verb|NP→UN : R isNP₌ → R isUN|\\
    \verb|ii : R isWN × R isUN → R isCR|\\
    \verb||\texttt{UN}$^{\to}$ \verb|∧WN→CR : R |\texttt{isUN}$^{\to}$ \verb| → R isWN → R isCR|\\
    \verb|iii : subcommutative R → R isCR|\\
\end{proof}
\begin{proof}[Discussion.]
    In $\mathtt{UN^{→}∧WN→CR}$ a more general proof of ($ii$) is given using $\UNto$.
\end{proof}
\apcomment{Perhaps change the fourth implication to use $\times$ instead of $\to$ ??}

>>>>

The following is a summary of the differences in definitions we have used as mentioned in Section \ref{sec:Definitions}.
\begin{itemize}
    \item $\SN$ : We define this via the notion of accessibility rather than the non-existence of infinite reduction sequences. Both definitions articulate normalization as well-foundedness of the converse relation. However, different notions of well-foundedness are used in the two definitions. The differences are highlighted in Section~\ref{sec:Well-foundedness}.
     See Section \ref{sec:Well-foundedness} for further discussion.
    \item $\NPe$ : We use the equivalent $\NP$ definition.
    \item $\Inc$ : We use the more general property $\RP$.
\end{itemize}

As mentioned, the $\Inc$ property can impose a computationally demanding requirement. However, $\Inc$ is used only in Theorem 1.2.3
and the only crucial aspect of $\Inc$ needed for the proof is that no $R$-increasing
sequence can have an upper bound.
We establish that the weaker property $\RP$ (or its equivalent $\RPm$) is sufficient for replacing $\Inc$ in the theorem (see Subsection \ref{subsec:theorems}).
This led us to examine the condition on $s (i)$ appearing inside $\RP$ more closely. This condition provides a generalization of the notion of a normal form,
which we explore in detail in Section \ref{sec:Hierarchy}.

@@

\begin{proof}
    See \texttt{module Theorem-1-2-2} in \texttt{Theorems.agda}, \\
    \verb|CR→NP : R isCR → R isNP₌|\\
    \verb|NP→UN : R isNP₌ → R isUN|\\
    \verb|ii     : R isWN × R isUN → R isCR|\\
    \verb|ii+    : R isWN × R |\texttt{isUN}$^{\to}$ \verb|→ R isCR|\\
    \verb|iii    : subcommutative R → R isCR|\\
\end{proof}
\begin{proof}[Discussion.]
    In $\mathtt{UN^{→}∧WN→CR}$ a more general proof of ($ii$) is given using $\UNto$.
\end{proof}

### 2025-07-08 Relations/Paper/Implications.tex
Implications and summertodo minor changes update
Changes made on meeting 8th July

% From the perspective of programming language theory, the two properties of abstract rewriting
% systems that are most interesting are $\SN$ and $\CR$. The following hierarchies
% show chains of implications starting from these properties.
This section examines the hierarchy of ARS properties.
In programming language theory, $\SN$ and $\CR$ constitute the essential properties that
jointly establish completeness.
Consequently, our analysis centers on two primary categories of properties: those associated
with termination and those associated with confluence.
% Possibly redundant introductory paragraph

@@

\begin{definition}\label{def:mf} Let $a \in A$. \hfill
    \begin{description}
        \item[\ule{$a \in \MF_R$}] \emph{$a$ is a minimal form} if $a \mstep b \implies b \mstep a$ for any $b \in A$.
        \item[\ule{$a \in \WM_R$}] \emph{$a$ is weakly minimalizing} if $a \mstep b$ for some $b \in \MF_R$.
        \item[\ule{$a \in \SMseq_R$}] \emph{$a$ is sequentially strongly minimalizing} if for every $R$-increasing sequence $s$ starting at $a$, there exists
        an $i \in \nat$ such that $s (i) \in \MF$.
        \item[\ule{$a \in \SM_R$}]  \emph{$a$ is strongly minimalizing} if either $a \in MF_R$, or every element one $R$ step from $a$ is strongly minimalizing. (This
        definition is to be understood inductively).
        \item[\ule{$a \in \MP_R$}] \emph{$a$ has the minimal form property} if $c \bmstep a \mstep b \implies c \mstep b$ for any $b \in \MF$.
    \end{description}
\end{definition}

Trivially, the following holds.
\begin{proposition}\label{prop:nftomf}
    $\NF \implies \MF$
\end{proposition}
\begin{proof}
    See, \verb|NF ⊆ MF : ∀x → NFx → MFx| in \texttt{Implications.agda}.
\end{proof}

$\SMseq$ is only classically equivalent to $\SM$. The
following proposition relates these new definitions with our previous $\RP$ definitions.
\begin{proposition}\label{prop:SMRP}
    $SMseq \leftrightarrow RP \land BP$
\end{proposition}
\begin{proof}
    See,\\
    \verb|RP∧BP→SMseq : R isRP → R isBP → ∀ {x : A} → SMseq R x| \\
    \verb|RisSMseq→RisRP : (∀ {x : A} → SMseq R x) → R isRP|\\
    \verb|RisSMseq→RisBP : (∀ {x : A} → SMseq R x) → R isBP| \\
    in \texttt{Implications.agda}.
\end{proof}

@@

>>>>

This section examines the hierarchy of ARS properties.
In programming language theory, $\SN$ and $\CR$ constitute the essential properties that
jointly establish completeness.
Consequently, our analysis centers on two primary categories of properties: those associated
with termination and those associated with confluence.

@@

\begin{definition}\label{def:mf} Let $a \in A$. \hfill
    \begin{description}
        \item[\ule{$a \in \MF_R$}] \emph{$a$ is a minimal form} if $a \mstep b \implies b \mstep a$ for any $b \in A$.
        \item[\ule{$a \in \WM_R$}] \emph{$a$ is weakly minimalizing} if $a \mstep b$ for some $b \in \MF_R$.
        \item[\ule{$a \in \SM_R$}]  \emph{$a$ is strongly minimalizing} if either $a \in MF_R$, or every element one $R$ step from $a$ is strongly minimalizing. (This
        definition is to be understood inductively).
        \item[\ule{$a \in \SMseq_R$}] \emph{$a$ is sequentially strongly minimalizing} if for every $R$-increasing sequence $s$ starting at $a$, there exists
        an $i \in \nat$ such that $s (i) \in \MF$.
        \item[\ule{$a \in \MP_R$}] \emph{$a$ has the minimal form property} if $c \bmstep a \mstep b \implies c \mstep b$ for any $b \in \MF$.
    \end{description}
\end{definition}

Trivially, the following holds.
\begin{proposition}\label{prop:nftomf}
    $\forall x.\, x\in \NF \implies x \in \MF$
\end{proposition}
\begin{proof}
    See, \verb|NF ⊆ MF : ∀x → NFx → MFx| in \texttt{Implications.agda}.
\end{proof}

$\SMseq$ is only classically equivalent to $\SM$. The
following proposition relates these new definitions with our previous $\RP$ definitions.
\begin{proposition}\label{prop:SMRP}
    $\gSMseq \iff \gRP \land \gBP$
\end{proposition}
\begin{proof}
    See \texttt{Implications.agda},\\
    \verb|RP∧BP→SMseq    : R isRP → R isBP → ∀ {x : A} → SMseq R x| \\
    \verb|RisSMseq→RisRP : (∀ {x : A} → SMseq R x) → R isRP|\\
    \verb|RisSMseq→RisBP : (∀ {x : A} → SMseq R x) → R isBP| \\
\end{proof}

@@

\begin{proposition}\label{prop:MFtoNF} \hfill
    \begin{enumerate}
        \item $\forall x.\, x\in \MF \land x \in \WN \iff x \in \NF$
        \item $\forall x.\, x\in \MF \land x \in \SN \iff x \in \NF$
    \end{enumerate}

\end{proposition}
\begin{proof}
    See \texttt{module Normalizing-Implications} in \texttt{Implications.agda},\\
    \verb|MF∧WN↔NF : ∀ {x} → MF x × WN x ↔ NF x|\\
    \verb|MF∧SN↔NF : ∀ {x} → MF x × SN x ↔ NF x| \qedhere
\end{proof}

In contrast to the above proposition, the inclusion $\SN \subset \SM \cap \WN $ is strict: in Counterexample~\ref{CE:4}, $a \in \SM \cap \WN$, but $a \notin \SN$. In the tables below, we therefore include a seperate column for the combination of $\SM$ and $\WN$.

### 2025-07-08 Relations/Paper/Summer_2025_TODO.tex
Changes made on meeting 8th July

    \item (Review 2) "l 361: Swap s(k+1) and s(k) " We should clarify the definition here otherwise people (evidently) get confused.
    \item (Review 2) "l 422: $P_x$ has no arguments?"

@@

    \item (Review 3) In line 270, isWFseq- is used in Section 3 but defined only in Section 5.
    \item (Review 3) In line 192, the authors state that (in TeReSe) Ind denotes BP. However, in line 186, BP is defined from an R-increasing sequence, whereas this statement is not required in the definition of Ind by TeReSe. How does this difference impact the formalizations and generalizations pointed out in Section 3.3.3? (Maybe we should explicitly address all difference with TeReSe in one location.)

@@

\section*{Week starting April 28th}
\begin{itemize}
  \item Below notes from meeting on April 25th.
  \begin{todolist}
    \item  Fix Wellfounded.agda after changes made during meeting.
    \item Carry on refactoring Wellfounded.agda once fixed, breaking it up if necessary.
    \item [\done] Think about how to organize the Git repo going forward. For example, could be good to have a parent folder with ARS and WF, with a private dir for paper
    \item (Research item) How to go from WFseq to WFacc - possible that the Kleene Tree is a solution.
    \item (Research item) Go deeper in the related works and see which are most relevant. Make a list of works and give summaries to be used in future papers.
    \item (Research item) Find a concrete application of generalized Newman's Lemma.
    \item (Research item) Use TeReSe to create a list of rewrite systems.
    \item (Research item) Work on the missing implicaitons from the paper. So far we believe this is just CR $\to$ CP in the ARS section.
    \item (Research item) Formalize a first order term rewriting system (I can't remember what the motivation for this was).
    \item (Research item) Carry out research on rewriting counterexamples like our own so we can try and put them into some sort of context. Have our counerexamples been shown before?
    \item (Research item) Do we want to reincorporate Knaster Tarski?
  \end{todolist}

\end{itemize}
\section*{Week starting 26th May}
\begin{todolist}
  \item Double check the counterexamples match in the table to the diagrams.
  \item [\done] Make clear in proof references to agda code how it should be read (especially the first one). Make clear the module as well. After doing once, the others don't need to be as explicit.
  \item [\done] find a seperate symbol for -> in the code sections in the paper for UN$\to$. Superscipt would be perfect. $UN^{\rightarrow}$
  \item [\done] Mention theorem 123 ii and iii are inherently classical. For iii we require strong decidability.
  \item [\done] Working on CR $\to$ CP. Low priority. \apcomment{This is mentioned as one of the research items above}
\end{todolist}

\section*{Week starting 2nd June}
\begin{todolist}
  \item [\done] Should we add quantification to each definition(I think quantification for $a$ is already implied)? E.g.\\
  \item [\ule{$a \in \WCR_R$}] \emph{$a$ is weakly Church-Rosser (or weakly confluent)} if $c \bstep a \rstep b \implies c \mstep d \bmstep b$.\\
   Transforms to
  \item[\ule{$a \in \WCR_R$}] \emph{$a$ is weakly Church-Rosser (or weakly confluent)} if $c \bstep a \rstep b \implies c \mstep d \bmstep b$ for all $b, c, d \in A$.
  \item [\done] For showing inc -> RP we're currently just saying "This is immediate". Is this sufficient? We did have a proof previously but I think we removed it for space reasons.
  \item [\done] Remove "desired implications" and similar organizational issues in the code (general review)
  \item [\done] Consider defining FB in terms of existing list library; especially in the rebooted repo,
  we shouldn't use manual List type just for this.
  \item [\done] Move all definitions and properties from Section 2.3 to Seq.agda, except Inc.
  \item [\done] Revise section 3.1
\end{todolist}
\section*{Week starting 9th June (some of these items taken from Dr. Polonsky's email sent June 6th)}
\begin{todolist}
  \item [\done] Thm. 19(iv), i don't like that the parens are italicized.  maybe it would be okay if instead of CR => CP we just write "the forward implication"
  \item [\done] Proof of (iii), line 296: do we really need () around SN?
  \item Page 9, second half (the hierarchy): we should add a proof that $\MF \cap \SN \iff \NF$, say this in the text, and note that this does NOT hold for $\SM$ and $\WN$.  Hence, the tables include a separate column for their combination.
  \item  In the par before 4.1 counterexamples, add that in the global table, the counterexample 3 satisfies the hypotheses globally.
(I actually wonder whether we need example 9 at all, if we have a counterexample to the global one.  Am I missing something?). \sacomment{I don't think we need to expand that CE:3 satisfies the global hypotheses. I'm hoping that we've done enough to show to the reader how to read the table at this point. If you disagree I'm happy to add it in though. As to why have local and global counterexamples, good point. Unless either of us can think of a reason we'll remove in the next meeting.}
    \item [\done] the * and dagger in the table should be explained in the caption, or in the text.
    \sacomment{I prefer to have this as a footnote. I have put it as a caption for now but I would revert it to what we had before.  }
\end{todolist}

@@

>>>>

    \item [\done] (Review 2) "l 361: Swap s(k+1) and s(k) " We should clarify the definition here otherwise people (evidently) get confused.
    \item [\done](Review 2) "l 422: $P_x$ has no arguments?"

@@

    \item [\done](Review 3) In line 270, isWFseq- is used in Section 3 but defined only in Section 5.
    \item [\done] (Review 3) In line 192, the authors state that (in TeReSe) Ind denotes BP. However, in line 186, BP is defined from an R-increasing sequence, whereas this statement is not required in the definition of Ind by TeReSe. How does this difference impact the formalizations and generalizations pointed out in Section 3.3.3? (Maybe we should explicitly address all difference with TeReSe in one location.)

@@

\section*{Week starting April 28th}
\begin{itemize}
  \item Below notes from meeting on April 25th.
  \begin{todolist}
    \item [\done] Fix Wellfounded.agda after changes made during meeting.
    \item [\done] Think about how to organize the Git repo going forward. For example, could be good to have a parent folder with ARS and WF, with a private dir for paper

\end{itemize}
\section*{Week starting 26th May}
\begin{todolist}
  \item [\done] Double check the counterexamples match in the table to the diagrams.
  \item [\done] Make clear in proof references to agda code how it should be read (especially the first one). Make clear the module as well. After doing once, the others don't need to be as explicit.
  \item [\done] find a seperate symbol for -> in the code sections in the paper for UN$\to$. Superscipt would be perfect. $UN^{\rightarrow}$
  \item [\done] Mention theorem 123 ii and iii are inherently classical. For iii we require strong decidability.
  \item [\done] Working on CR $\to$ CP. Low priority. \apcomment{This is mentioned as one of the research items above}
\end{todolist}

\section*{Week starting 2nd June}
\begin{todolist}
  \item [\done] Should we add quantification to each definition(I think quantification for $a$ is already implied)? E.g.\\
  \item [\ule{$a \in \WCR_R$}] \emph{$a$ is weakly Church-Rosser (or weakly confluent)} if $c \bstep a \rstep b \implies c \mstep d \bmstep b$.\\
   Transforms to
  \item[\ule{$a \in \WCR_R$}] \emph{$a$ is weakly Church-Rosser (or weakly confluent)} if $c \bstep a \rstep b \implies c \mstep d \bmstep b$ for all $b, c, d \in A$.
  \item [\done] For showing inc -> RP we're currently just saying "This is immediate". Is this sufficient? We did have a proof previously but I think we removed it for space reasons.
  \item [\done] Remove "desired implications" and similar organizational issues in the code (general review)
  \item [\done] Consider defining FB in terms of existing list library; especially in the rebooted repo,
  we shouldn't use manual List type just for this.
  \item [\done] Move all definitions and properties from Section 2.3 to Seq.agda, except Inc.
  \item [\done] Revise section 3.1
\end{todolist}
\section*{Week starting 9th June (some of these items taken from Dr. Polonsky's email sent June 6th)}
\begin{todolist}
  \item [\done] Thm. 19(iv), i don't like that the parens are italicized.  maybe it would be okay if instead of CR => CP we just write "the forward implication"
  \item [\done] Proof of (iii), line 296: do we really need () around SN?
  \item [\done] Page 9, second half (the hierarchy): we should add a proof that $\MF \cap \SN \iff \NF$, say this in the text, and note that this does NOT hold for $\SM$ and $\WN$.  Hence, the tables include a separate column for their combination.
  \item [\done] In the par before 4.1 counterexamples, add that in the global table, the counterexample 3 satisfies the hypotheses globally.
(I actually wonder whether we need example 9 at all, if we have a counterexample to the global one.  Am I missing something?). \sacomment{I don't think we need to expand that CE:3 satisfies the global hypotheses. I'm hoping that we've done enough to show to the reader how to read the table at this point. If you disagree I'm happy to add it in though. As to why have local and global counterexamples, good point. Unless either of us can think of a reason we'll remove in the next meeting.}
    \item [\done] the * and dagger in the table should be explained in the caption, or in the text.
    \sacomment{I prefer to have this as a footnote. I have put it as a caption for now but I would revert it to what we had before.  }
\end{todolist}

@@

\section*{July TODO}
\begin{todolist}
  \item Carry on refactoring Wellfounded.agda once fixed, breaking it up if necessary.
  \item To finish the paper: We should do a final walk through and make sure we haven't introduced errors with our edits. In particular, do all the counterexamples line up. Do all the references to different sections and pieces of code in Agda line up. 
  \item Do we remove all our local counterexamples in favour of global ones?
  \item Rework well-foundedness section 
  \item Rework conclusion 
  \item Rework introduction 
  \item Do we introduce a new section for discussing Agda? Talking about why we think there is an advantage to doing this work in Agda vs (say) Isabelle

\section*{Research items}
\begin{todolist}
  \item (Research item) Go deeper in the related works and see which are most relevant. Make a list of works and give summaries to be used in future papers. * Look into the literature and related works. There should be an Agda formulation of rewrite systems (check agda standard library stuff on relations). 
  \item (Research item) Find a concrete application of generalized Newman's Lemma. * Finding a term rewrite system where it's eassier to see that it is weakly cr and that every term is SM (as opposed to being CR).  
  \item (Research item) Formalize a first order term rewriting system (I can't remember what the motivation for this was). Use TeReSe to create a list of rewrite systems.
  \item (Research item) Carry out research on rewriting counterexamples like our own so we can try and put them into some sort of context. Have our counerexamples been shown before?
  \item (Research item) Work on the missing implicaitons from the paper. So far we believe this is just CR $\to$ CP in the ARS section. - Will probably be quite a bit of work (low priority)
  \item (Research item) Do we want to reincorporate Knaster Tarski? - Not this paper 
  \item (Research item) How to go from WFseq to WFacc - possible that the Kleene Tree is a solution. - Not this paper

### 2025-07-08 Relations/Paper/Well-foundedness.tex
Changes made on meeting 8th July

\newcommand{\isWFacc}{\mathtt{isWFacc}}
\newcommand{\isWFseq}{\mathtt{isWFseq}}
\newcommand{\isWFmin}{\mathtt{isWFmin}}
\newcommand{\isWFaccm}{\mathtt{isWFacc-}}
\newcommand{\isWFseqm}{\mathtt{isWFseq-}}
\newcommand{\isWFminm}{\mathtt{isWFmin-}}

\newcommand{\isMinDec}{\mathtt{isMinDec}}

@@

In \cite{Terese}, well-foundedness is defined as in the classical mathematics textbooks,
as the absence of an infinite $R$-decreasing sequence.
Translating this statement directly into type theory yields the formula
\[\tag{$\isWFseqm$} \lnot \big(\exists s : \nat \to A \forall k : \nat. (s(k+1),s(k)) \in R \big) \]
(The label $\isWFseqm$ refers to the name of the corresponding term in \texttt{Wellfoundedness.agda}.)
Constructively, this is not a useful hypothesis, as it does not provide any positive computational content and requires an explicit sequence to be defined.

A stronger formulation asserts that, for any sequence, there exists an element
in the sequence that does not reduce to the next one.  This yields the statement
\[\tag{$\isWFseq$} \forall s : \nat \to A \exists k : \nat. (s(k),s(k+1)) \notin R \]

@@

On the other hand, $\isWFmin$ trivially implies that each $x \in A$ is weakly normalizing.
Indeed, given $x \in A$, consider the predicate $P_x(y) = x \mstep y$.
This predicate is non-empty, since the empty reduction yields a proof of $P_x(x)$.
Let $n$ be a $P_x$-minimal element.  Then $x \mstep n$.  Moreover, for any $y$,
If $x \rstep y$, then $\lnot(P_x)$.  But since reductions from $x$ are closed downward,
no such $y$ can exist, and hence $n$ is a normal form of $x$.

>>>>

@@

In \cite{Terese}, well-foundedness is defined as in the classical mathematics textbooks,
as the absence of an infinite $R$-decreasing sequence.
(Again, well-foundedness of $R$ corresponds to strong normalization of its converse.)
Translating this statement directly into type theory yields the formula

\[\tag{$\isWFseqm$} \lnot \big(\exists s : \nat \to A \forall k : \nat. \  R \,(s\,(k+1)) \, (s\, k) \big) \]

(The label $\isWFseqm$ refers to the name of the corresponding term in \texttt{Wellfoundedness.agda}.)
Constructively, this is not a useful hypothesis, as it does not provide any positive computational content and requires an explicit sequence to be defined.

A stronger formulation asserts that, for any sequence, there exists an element
in the sequence that does not reduce to the next one.  This yields the statement
\[\tag{$\isWFseq$} \forall s : \nat \to A \exists k : \nat.\, \lnot \, R \,(s\,k)\,(s\,(k+1)) \]

@@

On the other hand, $\isWFmin$ trivially implies that each $x \in A$ is weakly normalizing.
Indeed, given $x \in A$, consider the predicate $P_x(y) = x \mstep y$.
This predicate is non-empty, since the empty reduction yields a proof of $P_x(x)$.
Let $n$ be a $P_x$-minimal element.  Then $x \mstep n$.  Moreover, for any $y$,
If $n \rstep y$, then $\lnot(P_x(y))$.  But since reductions from $x$ are closed downward,
no such $y$ can exist, and hence $n$ is a normal form of $x$.
